---
author: "Zihao Liu and George Reynolds"
title: "EDAV PSet 2"
output: html_document
---
```{r, include=FALSE}
SOLUTIONS <- TRUE
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = SOLUTIONS,
                      eval = SOLUTIONS, cache = FALSE)
library(tidyverse)
library(rvest)
library(robotstxt)
library(stringr)
library(gardenR)
knitr::opts_chunk$set(echo = TRUE)
```


Read *Graphical Data Analysis with R*, Ch. 4, 5

Grading is based both on your graphs and verbal explanations. Follow all best practices as discussed in class.

```{r setup, include=FALSE}
# keep this chunk in your .Rmd file
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


### 1. Jungle Garden
Data: `garden_planting` from **gardenR** package

```
# install.packages("remotes")
# remotes::install_github("llendway/gardenR")
```

For each of the following choose the appropriate data from the package and draw a bar chart, following best practices, that can directly answer the following questions. For each, explain your choice of bar order.

a) How many varieties of each type of vegetable were planted?

```{r}
by_veg <- group_by(garden_planting, vegetable)
by_veg_var <- summarise(by_veg, varieties = n_distinct(variety))

ggplot(by_veg_var, aes(varieties, fct_reorder(`vegetable`, varieties))) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = seq(1, 13, 1)) +
  ggtitle("Varieties of Vegetables") + ylab("")
```

b) Were there any days of the week that Dr. Lendway never planted? Which days of the week was she most likely to be found planting? (Measure by the number of rows in `garden_planting`.)

```{r}
gp <- garden_planting
gp$dayofweek <- weekdays(as.Date(gp$date))

by_day <- group_by(gp, dayofweek)

by_day_freq <- by_day %>% count()
names(by_day_freq)[names(by_day_freq) == "n"] <- "observations"

by_day_freq %>% mutate(dayofweek=fct_explicit_na(dayofweek,"NA"))%>%
ggplot( aes(x = observations, y = fct_reorder(`dayofweek`, observations))) +
  geom_bar(stat = "identity") +
  ggtitle("Plantings by Day of Week") + ylab("")
```

Dr. Lendway was most likely to plant on Wednesdays and Saturdays. It seems that Dr. Lendway never planted on a Tuesday; however, it must be noted that there was one observation with no date recorded. It is possible that this NA value was in fact a Tuesday. Otherwise, this was never a planting day.

c) How many seeds were planted in each plot? 
```{r}
gp %>%
  group_by(plot) %>%
  summarise(seeds_planted = sum(number_seeds_planted, na.rm = T))  %>%
  arrange(seeds_planted) %>%
  mutate(plot=fct_reorder(plot,seeds_planted))%>%
    ggplot(aes(x=plot,y=seeds_planted))+
         geom_col()+
         coord_flip()+
       theme_grey()
```

I sort plots in descending order according to the numbers of seeds in.

### 2. Vaccination Rates

To get the data for this problem, we’ll scrape data from this page: https://www.beckershospitalreview.com/public-health/states-ranked-by-percentage-of-covid-19-vaccines-administered.html. Important: you should only execute parts (a) and (b) once. After running these lines, comment them out, and leave them commented out in the submitted version.

a) Check that it's ok to scrape the site. What is the result?

``` {r}
# robotstxt::paths_allowed("https://www.beckershospitalreview.com/public-health/states-ranked-by-percentage-of-covid-19-vaccines-administered.html") 

```

The check on the url returned "TRUE".

b) Read the page at the URL listed above and save it.

```{r}
# vaxes <- "https://www.beckershospitalreview.com/public-health/states-ranked-by-percentage-of-covid-19-vaccines-administered.html"
# vaxes <- read_html(vaxes)
vaxes <- read_html("vaccines.html")
```

c) Read in the page saved in part b) and create a data frame with the following columns: `State`, `Distributed`, `Administered`, `Percentage`. Display the first six rows of the data frame.

```{r}
list_items <- vaxes %>%
  html_node("body") %>%
  html_nodes("#content") %>%
  html_nodes("#left-column") %>%
  html_nodes(".full-article-wrapper") %>%
  html_nodes("ol") %>%
  html_nodes("li") 

df <- data.frame(State = character(),
                 Distributed = character(),
                 Administered = character(),
                 Percentage = double())

state_name <- function(state) {
  state_name <- word(list_items[[state]] %>% html_text()
                   , sep = "\\Doses")
  state_name <- str_remove(state_name, "\n")
  return(state_name)}

distributed <- function(state){
  distributed <- word(
    word(list_items[[state]] %>% html_text(), 2, 
         sep = "\\Doses"), 
    5)
  return(distributed)
}

administered <- function(state){
  administered <- trimws(sapply(str_split(word(word(list_items[[state]] %>% html_text(), 3, sep = "\\:"), sep = "\\of"), "P"), `[`,1))

  return(administered)
}

for (state in 1:50){
  df[nrow(df) + 1,] <- c(state_name(state)
                    , distributed(state)
                    , administered(state)
                    , 0 )}

df$Distributed <- as.numeric(gsub(",", "", df$Distributed))
df$Administered <- as.numeric(gsub(",", "", df$Administered))
df$Percentage <- df$Administered/df$Distributed

head(df)
```

d) Create a Cleveland dot plot showing the percentage of distributed vaccines that have been administered.

```{r fig.height=20}
ggplot(df, aes(Percentage, fct_reorder(`State`, Percentage))) +
  geom_point() +
  ggtitle("Percentage of Distributed Doses Administered") + ylab("")
```

e) Create a multiple Cleveland dot plot showing doses distributed in one color and doses administered in another for each state. Experiment with different sorting orders. Choose the one that you think works best and explain why.

```{r fig.height = 20}
ggplot(df, aes(Distributed, fct_reorder(`State`, Distributed), colour = "Distributed")) +
  geom_point() +
  geom_point(aes(x = Administered, y = State, colour = "Administered")) +
    scale_colour_discrete(name = "Statistic") +
  ggtitle("Doses Distributed vs. Administered") +
  ylab("") + xlab("Doses")
  

```

I chose to order the Cleveland dot plot by doses distributed. The other options I considered were doses administered and percentage of doses distributed administered. I chose doses distributed because I think that the most important use for this plot would be to get a sense of state efficacy in vaccine rollout. In this light, doses distributed represents a sample size, and administered the portion of our sample taking on the value of 1 for an indicator variable that demonstrates whether a distributed dose was administered. In this way, doses administered would also be very helpful to see the graph ordered by, but distributed represents the sample size and, therefore, how confident we can be that the results are truly reflective of the state's efficacy - for example, this choice shows that while WV was at the top of the percentage plot, this may not hold as it is largely driven by a small number of doses distributed. This seems to be the correct order to consider the states in for my purposes. (It is also worth noting that the different choices for order do not make significant differences at the top of the plot, the area most people will naturally consider first).

### 3. Taxis

Data: NYC yellow cab rides in June 2020, available here:  

https://www1.nyc.gov/site/tlc/about/tlc-trip-record-data.page

It's a large file so you may work with a reasonably-sized random sample of the data (for example 10000 rows). If you remove outliers / impossibilities explain your logic.

Draw four plots of `tip_amount` vs. `fare_amount` with the following variations:

(a) Scatterplot with alpha blending
```{r}
data3 <- read_csv("yellow_tripdata_2020-06.csv") %>% na.omit()
set.seed(1234)
data3_1 <- data3[sample(1:nrow(data3), 10000), ]

data3_a <- data3_1 %>%
filter(between(fare_amount, 0, quantile(fare_amount, 0.75)+3*IQR(fare_amount)) &
between(tip_amount, 0,  quantile(tip_amount, 0.75)+3*IQR(tip_amount)))

ggplot(data3_a, aes(fare_amount, tip_amount)) +
  geom_point(alpha = 0.5, size = 0.2) +
  theme_bw()
```

By observing raw data, we can find some features like fare_amount and tip_amount have a very large range of values. There are many ways to locate the outliers in a dataset. I choose to use interquartile range to select the outliers. An outlier would be a point below [Q1- (1.5)IQR] or above [Q3+(1.5)IQR] while values above [Q3 + 3xIQR] or below [Q1 - 3xIQR] are considered as extreme points. In that, Q1= 25th percentile, Q3= 75th percentile and IQR= Q3 – Q1. I tried both of them and found the former will remove so many points. So I choose to set the coefficient of IQR as 3 to make more values available for us and remove some extreme outliers at the same time. In that case, I observed that the lower bound, is negative which doesn't match the reality. Therefore I modified the lower bound to zero to make it more reasonable. 

(b) Scatterplot with alpha blending + density estimate contour lines
```{r}
ggplot(data3_a, aes(fare_amount, tip_amount)) +
  geom_point(alpha = 0.5,size = 0.2) +
  theme_bw() +
  geom_density_2d(alpha = 0.5)
```

(c) Hexagonal heatmap of bin counts

```{r}
ggplot(data3_a, aes(fare_amount, tip_amount)) +
  geom_hex(bins=30) +
  theme_bw()
```

(d) Square heatmap of bin counts 
```{r}
ggplot(data3_a, aes(fare_amount, tip_amount)) +
  geom_tile(stat = "binhex") +
  scale_fill_gradient(low="lightblue", high="yellow") +
  theme_bw()
```
For all, adjust parameters to the levels that provide the best views of the data.

(e) Describe noteworthy features of the data, using the "Movie ratings" example on page 82 (last page of Section 5.3) as a guide.  
```{r}
ggplot(data3_a, aes(trip_distance,tip_amount )) +
  geom_point(alpha=0.4) +
  xlim(0,15)+
  theme_bw()
```

From the plot, we can get some sights:

(1)Tips are proportional to distance and tips generally increase as distance increases.

(2)Passengers who took a long trip(more than 4 miles) always tip.

(3)Most tips are more than 1 dollar.

(4)When distance increases, people are more likely to tip in round figure.

(5)Most tips are concentrated in the interval of [1,5].